Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-10-18 paper.tex
rough restructuring

@@

@@

To introduce the $\findpath$ algorithm, we need the following definitions.
Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
We then say that $v$ \emph{induces} $C$.
\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
Note that the list of all
\todo{I added `non-trivial' here because the clusters corresponding to leaves cannot be put on the list.
Check that the indexing of $C_i$ is still correct.
I think we need to be explicit whether we ever use trivial clusters and I suggest that we don't.}
non-trivial
clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.

To compute a path from tree $T$ to $R$ we consider clusters $C_1, \ldots, C_{n-1}$ of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-1$.
\todo{In that case the number of clusters should be $n-2$.}
Trivial clusters do not need to be changed on any path between two trees which is why they are not considered in the algorithm.
First, set $\hat{T} = T$
Then for each $i$, find the node that induces $C_i$ in $\hat T$ and denote this node by $\mrca_{\hat{T}}(C_i)$ (short for \emph{most recent common ancestor}).
Update $\hat{T}$ by
\todo{This presumes  that the rank is higher.
Why?}
decreasing
the rank of $\mrca_{\hat{T}}(C_i)$ by performing $\rnni$ moves until it has the same rank $i$ as in $R$.
\todo{Do we need to explain why there is always an $\rnni$ move that moves the $\mrca$ down?
AG: We need to explain quite a bit here -- there  are many ways one can do those moves + comment above.}
Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}(C_i)$ is
\todo{What is the `parent of a cluster' -- please check that you follow definitions you introduce.}
the parent of cluster
$C_i$ in $\hat{T}$ after these $\rnni$ moves.
Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
\todo{This kind of structure is not ideal (see my comment on the paper structure above).
If we talk about the algorithm here, all the things, including the pseudo-code should be here.
Possible solution: have algorithmic stuff first and geometry second.
If you do it like this, explain that we need the algorithms for the geometry section.}
Further details, including pseudo-code for $\findpath$, can be found in Section~\ref{section:computation_findpath}.

This algorithm gives an upper bound for the diameter of $\rnni$, which
\todo{Why is that?}
improves
the one given in \autocite{Gavryushkin2018-ol}.
\todo{When}
Later we will see that this bound is tight.
However, we don't know how good $\findpath$ approximates distances in $\rnni$. \todo{Can we say something about the approximation ratio?}
More details can be found in Section~\ref{section:computation}.

@@

In this section we restrict our attention to the set of \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
We discovered this by computing caterpillar paths using an algorithm to compute shortest paths within the set of caterpillar trees and comparing them with shortest path in $\rnni$.
This algorithm will be explained in detail later in this section, after demonstrating why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.
Further details on our computational results can be found in Section~\ref{section:computation}.

@@

It is important to note that this algorithm can be applied to any input tree $T$, which does not need to be a caterpillar tree.
However, the results provided in this section are only true for caterpillar trees.
The pseudocode for $\mdtree$ can be found in Section~\ref{section:computation}.
\todo{figure? + do we need the pseudocode?}

@@

\section{Computational results}
\label{section:computation}

In \autocite{Gavryushkin2018-ol} an efficient algorithm for computing the $\rnni$ graph has been introduced.
We implemented this algorithm, which produces the graph for trees with up to $8$ taxa in reasonable time. \todo{Shall we say more about the algorithm here? + running time (in hours/minutes) on which computer?}
This enables us to compute exact distances between trees with up to $8$ taxa.
In particular, comparing exact distances with approximations as introduced in previous sections and verifying the Cluster Theorem of Section~\ref{section:split_theorem} can now be for small trees.
All algorithms have been implemented with Python and are available online. \todo{link?}
In the following, further details about the implementation of algorithms $\findpath$ and $\mdtree$ as well as a description of how to prove the Cluster Theorem will be given.

\subsection{Algorithm $\findpath$}
\label{section:computation_findpath}

In Section~\ref{section:geometry} we introduced the algorithm $\findpath$ for computing paths between arbitrary trees in $\rnni$.
Therefore, it approximates distances in $\rnni$.
In this section we are providing pseudocode for $\findpath$ (Algorithm~\ref{alg:find_path}) to provide further insights.
We implemented and tested this algorithm for trees on up to $k$ taxa, for which it gives the exact $\rnni$ distance. \todo{$k = ?$, How exactly did we test it?}
The running time of $\findpath$ is quadratic in the number of taxa.

\subsection{Algorithm $\mdtree$}
\label{section:computation_mdtree}

>>>>

\todo{LC: move the following definition? - it's actually quite important not only for the algorithm, so we might want to have it in the introduction?}
Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
We then say that $v$ \emph{induces} $C$.
\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
Note that the list of all
\todo{I added `non-trivial' here because the clusters corresponding to leaves cannot be put on the list.
Check that the indexing of $C_i$ is still correct.
I think we need to be explicit whether we ever use trivial clusters and I suggest that we don't.}
non-trivial
clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.

@@

\todo{LC: Explain structure of paper}

\section{Algorithms}
\label{section:algorithms}

\todo{LC: make sure that we move all definitions in this paper to the right place}

Within this section we will two introduce algorithms, namely $\findpath$ and $\csort$ for computing paths between two given trees in $\rnni$.
These will be used later to prove some geometric properties of the $\rnni$ graph.
Additionally we will present the algorithm $\mdtree$, which computes a caterpillar tree with maximum distance
\todo{LC: can we use the term max dist here?}
from a given tree.

% In \autocite{Gavryushkin2018-ol} an efficient algorithm for computing the $\rnni$ graph has been introduced.
% We implemented this algorithm, which produces the graph for trees with up to $8$ taxa in reasonable time. \todo{Shall we say more about the algorithm here? + running time (in hours/minutes) on which computer?}
% This enables us to compute exact distances between trees with up to $8$ taxa.
% In particular, comparing exact distances with approximations as introduced in previous sections and verifying the Cluster Theorem of Section~\ref{section:split_theorem} can now be for small trees.
% All algorithms have been implemented with Python and are available online. \todo{link?}
% In the following, further details about the implementation of algorithms $\findpath$ and $\mdtree$ as well as a description of how to prove the Cluster Theorem will be given.

\subsection{Algorithm $\findpath$}
\label{section:alg_findpath}

\todo{LC: Introducing words}

To compute a path from tree $T$ to $R$ we consider clusters $C_1, \ldots, C_{n-1}$ of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-1$.
\todo{In that case the number of clusters should be $n-2$.}
Trivial clusters do not need to be changed on any path between two trees which is why they are not considered in the algorithm.
First, set $\hat{T} = T$
Then for each $i$, find the node that induces $C_i$ in $\hat T$ and denote this node by $\mrca_{\hat{T}}(C_i)$ (short for \emph{most recent common ancestor}).
Update $\hat{T}$ by
\todo{This presumes  that the rank is higher.
Why?}
decreasing
the rank of $\mrca_{\hat{T}}(C_i)$ by performing $\rnni$ moves until it has the same rank $i$ as in $R$.
\todo{Do we need to explain why there is always an $\rnni$ move that moves the $\mrca$ down?
AG: We need to explain quite a bit here -- there  are many ways one can do those moves + comment above.}
Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}(C_i)$ is
\todo{What is the `parent of a cluster' -- please check that you follow definitions you introduce.}
the parent of cluster
$C_i$ in $\hat{T}$ after these $\rnni$ moves.
Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
\todo{This kind of structure is not ideal (see my comment on the paper structure above).
If we talk about the algorithm here, all the things, including the pseudo-code should be here.
Possible solution: have algorithmic stuff first and geometry second.
If you do it like this, explain that we need the algorithms for the geometry section.}

\todo{LC: Change the references to sections}
In Section~\ref{section:geometry} we introduced the algorithm $\findpath$ for computing paths between arbitrary trees in $\rnni$.
Therefore, it approximates distances in $\rnni$.
In this section we are providing pseudocode for $\findpath$ (Algorithm~\ref{alg:find_path}) to provide further insights.
We implemented and tested this algorithm for trees on up to $k$ taxa, for which it gives the exact $\rnni$ distance. \todo{$k = ?$, How exactly did we test it?}
The running time of $\findpath$ is quadratic in the number of taxa.

\subsection{Algorithm $\cpath$}
\label{section:alg_cpath}

\todo{LC: move definitions of caterpillar stuff here}
\todo{LC: pseudo code}

\subsection{Algorithm $\mdtree$}
\label{section:alg_mdtree}

@@

This algorithm gives an upper bound for the diameter of $\rnni$, which
\todo{Why is that?}
improves
the one given in \autocite{Gavryushkin2018-ol}.
\todo{When}
Later we will see that this bound is tight.
However, we don't know how good $\findpath$ approximates distances in $\rnni$. \todo{Can we say something about the approximation ratio?}

@@

In this section we restrict our attention to the set of \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
We discovered this by computing caterpillar paths using an algorithm to compute shortest paths within the set of caterpillar trees and comparing them with shortest path in $\rnni$.
This algorithm will be explained in detail later in this section, after demonstrating why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.

@@

It is important to note that this algorithm can be applied to any input tree $T$, which does not need to be a caterpillar tree.
However, the results provided in this section are only true for caterpillar trees.
The pseudocode for $\mdtree$ can be found in Section~\ref{section:alg_mdtree}.
\todo{figure? + do we need the pseudocode?}

@@

\todo{LC: connect previous with following (has been in computation section before)}
